---
title: "Project7Warshak"
author: "Jesse Warshak"
date: "2024-03-17"
output: html_document
---
```{r}
library(tidyverse)
library(ggplot2)
```


```{r}
collegedata <- read.csv("college.csv")
numvar <- ncol(collegedata)
numobv <- nrow(collegedata)
missingv <- any(is.na(collegedata))

numvar
numobv
missingv
```

```{r}
collegedata %>%
  group_by(state)%>%
  ggplot()+
  geom_bar(aes(x = admission_rate, fill = state))
```


```{r}
collegedata %>%
  ggplot()+
  geom_point(aes(x = median_debt, y = loan_default_rate))+
  geom_smooth(aes(x = median_debt, y = loan_default_rate), method = "gam")
```

```{r}
ggplot(collegedata, aes(x = median_debt, y = loan_default_rate, size = undergrads)) +
  geom_point(alpha = 0.7) + 

geom_smooth(method = "loess", formula = y ~ x, aes(weight = undergrads), se = FALSE)
```


```{r}
ggplot(collegedata, aes(x = undergrads, fill = highest_degree)) +
  geom_density(alpha = 0.5) +
  labs(x = "Number of Undergraduates", y = "Density",
       title = "Distribution of Undergraduates by Highest Degree Offering") +
  scale_fill_discrete(name = "Highest Degree")
```
The number of undergraduates increases when the highest degree level increases from associate to bachelor to graduate. the number of undergraduates is very low when the highest level is associates.

```{r}
collegedata$log_num_undergrads <- log10(collegedata$undergrads)
ggplot(collegedata, aes(x = log_num_undergrads, fill = highest_degree)) +
  geom_density(alpha = 0.5) +
  labs(x = "Log10(Number of Undergraduates)", y = "Density",
       title = "Distribution of Log-Transformed Undergraduates by Highest Degree Offering") +
  scale_fill_discrete(name = "Highest Degree")
```
I believe this shows the date better.


```{r}
collegedata$log_num_undergrads <- log10(collegedata$undergrads)
ggplot(collegedata, aes(x = log_num_undergrads, fill = highest_degree)) +
  geom_density(alpha = 0.5, position = "stack") +
  labs(x = "Log10(Number of Undergraduates)", y = "Density",
       title = "Composition of Undergraduates by Highest Degree Offering") +
  scale_fill_discrete(name = "Highest Degree")
```
graduate and bachelor seem to be the highest proportion for highest degree type. graduate being the highest.
```{r}
degree_counts <- table(collegedata$highest_degree)

barplot(degree_counts, 
        main = "Number of Schools by Highest Degree Type",
        xlab = "Highest Degree Type",
        ylab = "Number of Schools",
        col = "blue", 
        border = "black",
        ylim = c(0, max(degree_counts) + 10)) 

text(x = barplot(degree_counts), y = degree_counts + 2, labels = degree_counts, pos = 3)
```

```{r}
ggplot(collegedata, aes(x = sat_avg, y = admission_rate, color = control)) +
  geom_point(alpha = 0.7) +
  labs(x = "Average SAT Score", y = "Admission Rate",
       title = "Association between Average SAT Score and Admission Rate") +
  scale_color_manual(values = c("public" = "blue", "private" = "orange"))
```

1. How does the gender impact the tuition?
```{r}
collegedata%>%
  ggplot()+
  geom_density_ridges_gradient(aes(x = tuition, y = gender))
```

2. Is the faculty salary higher in private or public schools?
```{r}
collegedata%>%
  ggplot()+
  geom_density_ridges_gradient(aes(x = faculty_salary_avg, y = control))
```
